### 2024-04-23 parts/transducers.tex
move to appendix and clean up

\AP
Given a function $f$, our goal is to leverage $\resleq{f}{k}$ to build a
canonical \kl{$\NPoly[k-1]$-transducer} that \kl{computes} $f$. The naïve
approach is to consider as states some representatives for the $\resleq{f}{k}$
minimal elements of $\Sigma^*$ (assuming there are finitely many of them), all
computing the same function (\cref{good-residual-ordering:fact}), and define
transitions by letting $\delta(u, a)$ be \emph{one} state $v$ such that $v
\resleq{f}{k} ua$. This naïve approach does not yield a canonical model, as
illustrated by the two distinct \kl{$\NPoly[0]$-transducers} of
\cref{non-canonical-transd:fig} computing the function $\BadExOk$ of
\cref{non-canonical-transd:ex}, and having as states minimal elements for $\resleq{\BadExOk}{0}$.
To obtain a
canonical notion, we enforce that the set of states is a \intro{downwards
closed} subset of $\Sigma^*$ for the prefix ordering $(\prefleq)$, that is, if
$u \in \Sigma^*$ is a state, then every prefix of $u$ must also be a state.

\begin{definition}
    \label{residual-transducer:def}
    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
    A transducer $\aTransd \defined (Q, q_0, \delta, \lambda, F)$
    is a \intro{$k$-residual transducer}
    of $f$ 
    when
    it is a \kl{$\NPoly[k-1]$-transducer}
    satisfying the following properties:
    \begin{enumerate}
        \item $\aTransd$ \kl{computes} $f$;
        \item $Q \subseteq \Sigma^*$ is a \kl{downwards closed}
            for $\prefleq$;
        \item $q_0 = \varepsilon$;
        \item every state $q \in Q$ is accessible from $q_0$;
        \item For all $u, a \in Q$,
            $\delta(u,a)$ is the $\prefleq$-maximal $v \in Q$
            such that $v \prefleq ua$, and $v \resleq{f}{k} ua$.
        \item For all $u,a \in Q$,
            $\lambda(u,a) = \app{f}{ua} - \app{f}{\delta(u,a)}$.
    \end{enumerate}
\end{definition}

Because of the restriction on prefixes, the \kl{$k$-residual transducer} is
actually unique when it exists. We now introduce \cref{residual:algo} to
compute a \kl{$k$-residual transducer} given a function $f$. Notice that this
algorithm requires the ability to test if a function belongs to $\NPoly[k]$,
which is only known to be feasible for \kl{commutative} \kl{polyregular
functions}. However, the termination of this algorithm also proves the
existence of the \kl{$k$-residual transducer}.

\begin{algorithm}
    $Q \defined \{ \varepsilon \}$;
    $O \defined \setof{ a }{ a \in \Sigma}$;
    $\delta \defined \emptyset$;
    $\lambda \defined \emptyset$;
    $F \defined \emptyset$;

@@

The following \cref{n-poly-k-implies-wqo:lemma} is the core argument of the
section, and states that for a function $f \in \NPoly[k]$, not only $(\prefleq
\Rightarrow \resleq{f}{k})$ is \kl{well}, but the quasi-ordering $(\resleq{f}{k})$ is a
\kl{well-quasi-ordering}. This has to be compared with the property of being
\kl{$(k,\Nat)$-combinatorial}, which states that ultimately, the function is
increasing in every direction.

\begin{theorem}
    \label{non-commutative-npoly:thm}
    Let $f \in \ZPoly$ be a non-negative function, 
    and $k \in \Nat$,
    the following are equivalent:
    \begin{enumerate}
        \item \label{n-poly-1-transd:item} $f$ is \kl{computed}
            by a \kl{$\NPoly[k-1]$-transducer};
        \item \label{n-poly-k:item} $f \in \NPoly[k]$;
        \item \label{n-poly-wqo:item} $(\Sigma^*, \resleq{f}{k})$ is a
            \kl{well-quasi-ordering};
        \item \label{n-poly-well:item} every $\prefleq$-increasing sequence
            of $\Sigma^*$  is a \kl{good sequence}
            for $\resleq{f}{k}$;
        \item \label{n-poly-residual:item} $f$ is \kl{computed} by a
            \kl{$k$-residual transducer}.
    \end{enumerate}
    If $f$ is assumed to be \kl{commutative}, then all the above
    properties are decidable, and the conversions effective.
\end{theorem}

\AP
We can prove that for functions in $\NPoly[0]$, aperiodicity is equivalent to
the absence of counters in the \kl{residual transducer} of the function (see
\cref{aperiodic-iff-residual:lem}), however this result does not generalize to
higher growth rates, as witnessed by \cref{non-aperiodic-residual-transd:ex}.
In the example, no \kl{$\NPoly[0]$-transducer} with the minimal number of
states computing the function is \kl{counter-free}, which suggests that a more
sophisticated object is needed to observe aperiodicity. In the spirit of the
characterization of $\NPoly$ based the $(\resleq{f}{k})$ quasi-ordering
(\cref{non-commutative-npoly:thm}), we introduce the notion of \intro{aperiodic
ordering} of $\Sigma^*$ as follows: $(\Sigma^*, \leq)$ is \reintro[aperiodic
ordering]{aperiodic} whenever for all $u, w \in \Sigma^*$, there exists $N_0
\in \Nat$, such that the sequence $\seqof{uw^n}{n \geq N_0}$ is increasing.
This leads to the following \cref{sf-no-periods-on-sequences:conj}, of which
the easy implication is already known to hold
(\cref{sf-no-periods-on-sequences:lemma}), and requires the introduction of the
\emph{star-free variant} $(\intro*\resleqsf{f}{k})$ of $(\resleq{f}{k})$,
defined by $u \reintro*\resleqsf{f}{k} v$ whenever $(\app{f}{v} - \app{f}{u})
\in \NSF[k-1]$.

@@

>>>>

\AP Given a function $f$, our goal is to leverage $\resleq{f}{k}$ to build a
canonical \kl{$\NPoly[k-1]$-transducer} that \kl{computes} $f$. The naïve
approach is to consider as states some representatives for the $\resleq{f}{k}$
minimal elements of $\Sigma^*$ (assuming there are finitely many of them),
%all computing the same function (\cref{good-residual-ordering:fact}),
and define transitions by letting $\delta(u, a)$ be \emph{one} state $v$ such
that $v \resleq{f}{k} ua$. This naïve approach does not yield a canonical
model, as illustrated by the two distinct \kl{$\NPoly[0]$-transducers} of
\cref{non-canonical-transd:fig} computing the function $\BadExOk$ of
\cref{non-canonical-transd:ex}, and having as states minimal elements for
$\resleq{\BadExOk}{0}$. To obtain a canonical notion, we enforce that the set
of states is a \intro{downwards closed} subset of $\Sigma^*$ for the prefix
ordering $(\prefleq)$, that is, every prefix of a state is also a state.

\begin{definition}
    \label{residual-transducer:def}
    Let $f \colon \Sigma^* \to \Nat$ and $k \in \Nat$.
    A transducer $\aTransd \defined (Q, q_0, \delta, \lambda, F)$
    is a \intro{$k$-residual transducer}
    of $f$ 
    when
    it is a \kl{$\NPoly[k-1]$-transducer}
    satisfying the following properties:
    \begin{enumerate}
        \item $\aTransd$ \kl{computes} $f$;
        \item $Q \subseteq \Sigma^*$ is a \kl{downwards closed}
            for $\prefleq$;
        \item $q_0 = \varepsilon$;
        \item every state $q \in Q$ is accessible from $q_0$;
        \item For all $u, a \in Q$,
            $\delta(u,a)$ is the $\prefleq$-maximal $v \in Q$
            such that $v \prefleq ua$, and $v \resleq{f}{k} ua$.
        \item For all $u,a \in Q$,
            $\lambda(u,a) = (\app{f}{ua} - \app{f}{\delta(u,a)}) \in \NPoly[k-1]$.
    \end{enumerate}
\end{definition}

Because of the restriction on prefixes, the \kl{$k$-residual transducer} is
actually unique when it exists. We now introduce \cref{residual:algo} to
compute the \kl{$k$-residual transducer} given a function $f$. Notice that this
algorithm requires the ability to test if a function belongs to $\NPoly[k]$,
which is only known to be feasible for \kl{commutative} \kl{polyregular
functions}. However, the termination of this algorithm also proves the
existence of the \kl{$k$-residual transducer}. The key argument proving the
termination of \cref{residual:algo} is based on the fact that for a function $f
\in \NPoly[k]$, the quasi-ordering $(\resleq{f}{k})$ is a
\kl{well-quasi-ordering}.

\begin{algorithm}[t]
    $Q \defined \{ \varepsilon \}$;
    $O \defined \setof{ a }{ a \in \Sigma}$;
    $\delta \defined \emptyset$;
    $\lambda \defined \emptyset$;
    $F \defined \emptyset$;

@@

\begin{theorem}
    \label{non-commutative-npoly:thm}
    Let $f \in \ZPoly$ be a non-negative function, 
    and $k \in \Nat$,
    the following are equivalent:
    \begin{enumerate}
        \item \label{n-poly-1-transd:item} $f$ is \kl{computed}
            by a \kl{$\NPoly[k-1]$-transducer};
        \item \label{n-poly-k:item} $f \in \NPoly[k]$;
        \item \label{n-poly-wqo:item} $(\Sigma^*, \resleq{f}{k})$ is a
            \kl{well-quasi-ordering};
        \item \label{n-poly-well:item} every $\prefleq$-increasing sequence
            of $\Sigma^*$  is a \kl{good sequence}
            for $\resleq{f}{k}$;
        \item \label{n-poly-residual:item} $f$ is \kl{computed} by a
            \kl{$k$-residual transducer}.
    \end{enumerate}
    If $f$ is assumed to be \kl{commutative}, the  
    properties are decidable, and the conversions effective.
    \proofref{non-commutative-npoly:thm}
\end{theorem}

\AP Unfortunately, the \kl{residual transducer} does not seem to provide an
algorithm to decide $\NSF$ inside $\NPoly$ in the general setting. While we can
prove that for functions in $\NPoly[0]$, aperiodicity is equivalent to the
absence of counters in the \kl{residual transducer} of the function (see
\cref{aperiodic-iff-residual:lem}), this result does not generalize to higher
growth rates, as witnessed by \cref{non-aperiodic-residual-transd:ex}.

@@

In the example, no \kl{$\NPoly[0]$-transducer} with the minimal number of
states computing the function is \kl{counter-free}, which suggests that a more
sophisticated object is needed to observe aperiodicity. In the spirit of the
characterization of $\NPoly$ based the $(\resleq{f}{k})$ quasi-ordering
(\cref{non-commutative-npoly:thm}), we introduce the notion of \intro{aperiodic
ordering} of $\Sigma^*$ as follows: $(\Sigma^*, \leq)$ is \reintro[aperiodic
ordering]{aperiodic} whenever for all $u, w \in \Sigma^*$, there exists $N_0
\in \Nat$, such that the sequence $\seqof{uw^n}{n \geq N_0}$ is increasing.
This leads to the following \cref{sf-no-periods-on-sequences:conj}, of which
the easy implication is already known to hold
(\cref{sf-no-periods-on-sequences:lemma}), and requires the introduction of the
\emph{star-free variant} $(\intro*\resleqsf{f}{k})$ of $(\resleq{f}{k})$,
defined by $u \reintro*\resleqsf{f}{k} v$ whenever $(\app{f}{v} - \app{f}{u})
\in \NSF[k-1]$.
